---
title: "How to write your paper in Rmd"
author: "Sau-Chin Chen"
date: "2019/1/28"
output: pdf_document
bibliography: "Orientation Cross Languages.bib"
---

```{r setup, include = FALSE}
  knitr::opts_chunk$set(
	echo = FALSE,
	cache = FALSE
)
  library("kableExtra")
  library("dplyr")
  library("papaja")
```

Below text was from https://psyarxiv.com/t2pjv/

## Data Collection Plan

This project follows the protocols of multi-lab registered replication reports [e.g., @CheungRegisteredReplicationReport2016; @HaggerMultilabPreregisteredReplication2016a] and the workflow of Psychological Science Accelerator [see @moshontzPsychologicalScienceAccelerator2018, Figure 2]. Until the submission of this proposal, there have been `r read.csv("Lab-summary.csv") %>% pull(Number.of.Laboratories) %>% sum()` participating laboratories from `r read.csv("Lab-summary.csv") %>% pull(Language) %>% nlevels()` languages (See registered laboratories and expected sample size in and \autoref{tab:lab-info}). Each registered laboratory will upload their publicly accessible pre-registered plan on the OSF. The principal investigators and the network committees will use the plans to guide the preparation of both stimuli and scripts.

<!--You can add comments like this and they will not appear in the knitted document -->

```{r lab-info, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
tbl_note <- paste("Recruitment of the participating laboratories is going on while we submitted this proposal. Last updated:", format(Sys.Date(), "%Y/%m/%d"))

read.csv("Lab-summary.csv") %>%
  apa_table(caption = "Summary of registered participating laboratories in the function of primary languages and sample size.",
            col.names = c("Language","Number of Laboratories","Registered Sample Size"),
            note = tbl_note,
            escape = TRUE)
```

## Reference
